GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2025-07-04 writeUps/paper.tex
resubmitted version JREE

\documentclass[]{article}
\usepackage[letterpaper]{geometry}
\usepackage{xr}
\externaldocument{writeUps/appendix}
\RequirePackage{amsmath,amsfonts,amssymb,amsthm}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage[caption=false]{subfig}
\usepackage[doublespacing]{setspace}

@@

@@

\usepackage{endfloat}% To `separate' figures and tables from text if required

@@

%\begin{aug}
%  \author{
%  Adam C. Sales\thanks{CONTACT Adam C. Sales. Email: asales@wpi.edu}, Kirk P. Vanacore, and Erin R. Ottmar%\\
%  %\affil{Worcester Polytechnic Institute, Worcester, MA, USA}
% }

% \maketitle

@@

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. For instance, one component of an educational computer application is the availability of “bottom-out” hints that provide the answer. In evaluating a recent experimental evaluation against alternative programs without bottom-out hints, researchers may be interested in estimating separate average treatment effects for students who, if given the opportunity, would request bottom-out hints frequently, and for students who would not. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and check. In this paper, we introduce a new M-estimation principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We present a simulation study that demonstrates the novel method's greater robustness compared to popular alternatives and illustrate the method through two real-data analyses.
\end{abstract}
% \begin{keywords}
% Causal Inference; Principal Stratification; Educational Technology
% \end{keywords}

@@

These concerns have prompted the development of alternative approaches, including bounding \citep{bounding} and randomization inference \citep{nolen2011randomization}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute an unbiased estimate of principal effects.
Principal ignorability is a direct analogue of the strong ignorability assumption in causal observational studies \citep[e.g.][]{rosenbaum1983central}---that all covariates that predict both principal stratum membership and outcomes have been accounted for---in other words, the precise type of assumption that randomized trials are designed to avoid.
A separate set of results in \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.

This paper expands on those latter results by presenting a new principal effect estimator for the case of one-way noncompliance, in which principal stratum membership is observed in one treatment group but not the other (future work will extend our approach to general scenarios). Our estimator boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally, estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.

After the initial submission of this manuscript, we became aware of \citet{richardson2023estimating}, which suggests a similar, though not identical, estimator to ours, using G-estimation of a structural nested mean model \citep[e.g.][]{wallace2017r}.
Our development of \textsc{geepers} builds on \citet{richardson2023estimating} by emphasizing connections to both parametric mixture modeling and principal score analysis, and comparing \textsc{geepers} to those methods in a simulation study.
Moreover, \textsc{geepers}, which, in practice, is a combination of logistic regression and OLS, is more accessible to applied researchers while yielding very similar results.

This work advances the literature in three ways: first, it makes an explicit connection between principal score methods, M-estimation, and semi-parametric estimation of principal effects using auxiliary data. Second, it provides a general estimator that relies on two of the most familiar tools in the applied education researcher's toolbox: linear and logistic regression. The \textsc{geepers} method can, therefore, make principal stratification methods accessible for a much broader range of applied education researchers. Lastly, this work includes an extensive simulation study that compares \textsc{geepers} to parametric mixture modeling and principal score weighting under a range of favorable and challenging scenarios. The simulation demonstrates the strengths and weaknesses of each method, as well as the trade-offs between them.

@@

\begin{table}
    \caption{\label{tab:factor} Manipulated factors in simulation study}
  \centering
  %\rowcolors{1}{}{lightgray}
\begin{tabular}{r|ll}%{*{3}{c}}
  \hline
  Factor &\multicolumn{2}{l}{Levels}\\
  \hline
Sample size per condition $n$ &\multicolumn{2}{l}{$100^*,200^*,300^*,400^*,500,600^*,700^*,800^*,900^*,1000$}\\
$\alpha$ &\multicolumn{2}{l}{$\rd{0},0.2,0.3^\dagger,0.4^\dagger,0.5,0.6^\dagger,0.7^\dagger,0.8^\dagger,0.9^\dagger,1^\dagger$}\\
The distribution of $\epsilon_i$&\multicolumn{2}{l}{ Normal, 
\rd{Uniform}}\\
%$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
\multirow{4}{*}{Interactions in \eqref{eq:y-sim}}& None:  $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
& \rd{$\bm{x}:\st$}: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
& \rd{$\bm{x}:Z$}: $\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
& \rd{$\bm{x}:\st$ \& $\bm{x}:Z$}: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% between $\st$ and covariates & No, Yes\\
% Interactions between $Z$ and covariates& No, Yes\\
\hline
\multicolumn{3}{p{4.5in}}{\footnotesize Factor levels in red violate assumptions of \geepers, \pmm, or both. Factors fully crossed except where indicated.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $*$ Only tested with $\alpha=0.5$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $\dagger$ Only tested with $n=500$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}

@@

\subsection{Results}\label{sec:simResults}
In this section, we evaluate estimates of $\eff1$. Results for estimates of $\eff0$ and $\eff1-\eff0$, which are similar, can be found in the online appendix. 
In replication $b=1,\dots,500$, define the error of a given estimator $\heff1_b$ as $err_b=\heff1_b-\eff1$. The bias of $\heff1$, $\EE[\heff1-\eff1]$, is estimated as $\widehat{bias}=\sum_b err_b/500$, and the empirical standard error is defined as $\left\{\sum_b(\heff1_b-\overline{\heff1})^2/499\right\}^{1/2}$, the sample standard deviation of $\heff1$. We also report the empirical coverage of confidence intervals, the proportion of replications in which $\eff1$ is within the estimated confidence bounds of $\heff1$. Results on root-mean-squared-error %$\left(\widehat{bias}^2+SE_{emp}^2\right)^{1/2}$ 
$\left(\sum_b err_b^2/500\right)^{1/2}$
are in the online appendix.

@@

\begin{figure}
  \centering
  \includegraphics{../simFigs/biasSEbyB1.jpg}
  \caption{Empirical bias and standard error of M- and Mixture estimates of $\muc1$ as $\alpha$ varies from 0.1 to 1. There were 500 replications for each $n$. Other factors were held fixed at the noted values.}
  \label{fig:alpha}
\end{figure}

@@

Our next set of results compares the performance of \geepers, \pmm, and \psw when one or more assumptions of each method may be violated.

% \begin{figure}[!ht]
%   \centering
%   \includegraphics[width=5.5in,clip]{../simFigs/boxplotsNew.pdf}
%   \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
%   \label{fig:boxplots}
% \end{figure}

@@

@@

@@

Another arguable disadvantage of the observational study approach is that it overlooks the broader context of the experiment, focusing solely on whether bottom-out hints affect posttest scores rather than examining their role in the overall effectiveness of ASSISTments.
A second approach, mediation analysis, contextualizes the bottom-out effect within the larger experiment.
It, too, relies on the same non-confounding assumption as an observational study, and also suffers from an additional challenge: mediation analysis relies on a ``common support'' condition that ``the mediator state must not be a deterministic function of the treatment'' \citep[p. 225]{celli2022causal}.
However, bottom-out hints are unavailable in the other three arms of the study, violating this condition (\citealt{sales2021student} offers a solution, which we find unconvincing).

@@

@@

@@

\section{Discussion}\label{sec:discussion}
\textsc{geepers} is a straightforward approach to principal effect estimation under strong monotonicity, built on widely-used regression models, which is more robust---though sometimes less precise---than alternative approaches under a wide array of scenarios.

@@

@@

>>>>

\documentclass[]{article}
\usepackage[letterpaper]{geometry}
\usepackage{xr}
\externaldocument{appendix}
\RequirePackage{amsmath,amsfonts,amssymb,amsthm}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage[caption=false]{subfig}
\usepackage[doublespacing]{setspace}

@@

\usepackage{placeins}

@@

%\usepackage{endfloat}% To `separate' figures and tables from text if required

@@

%\begin{aug}
  \author{
  Adam C. Sales\thanks{CONTACT Adam C. Sales. Email: asales@wpi.edu}, Kirk P. Vanacore, and Erin R. Ottmar%\\
   }

%   \maketitle
\date{}

@@

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. For instance, one component of an educational computer application is the availability of “bottom-out” hints that provide the answer. In evaluating a recent experimental evaluation against alternative programs without bottom-out hints, researchers may be interested in estimating separate average treatment effects for students who, if given the opportunity, would request bottom-out hints frequently, and for students who would not. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and check, making them inaccessible to many applied researchers. In this paper, we introduce a new principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We present a simulation study that demonstrates the novel method's greater robustness compared to popular alternatives and illustrate the method through two real-data analyses.
\end{abstract}
% \begin{keywords}
% Causal Inference; Principal Stratification; Educational Technology
% \end{keywords}

@@

Parametric mixture models are often difficult for applied researchers to implement---it is far removed from the regression models that dominate quantiative social science.
Fitting them may require familiarity with Bayesian Markov Chain Monte Carlo methods or liklihood optimization routines---even when these are automated \citep[e.g.]{PStrata}, they are slow and require an additional round of diagnostics for model convergence.
Furthermore, the usual model checking techniques of regression analysis, such as examining residuals, are not straightforward to apply in these contexts.
Combined with their fragility and heavy dependence on model specification, these factors have severely limited their accessibility to applied researchers.

These concerns have prompted the development of alternative approaches, including bounding \citep{bounding} and randomization inference \citep{nolen2011randomization}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute an unbiased estimate of principal effects.
Principal ignorability is a direct analogue of the strong ignorability assumption in causal observational studies \citep[e.g.][]{rosenbaum1983central}---that all covariates that predict both principal stratum membership and outcomes have been accounted for---in other words, the precise type of assumption that randomized trials are designed to avoid.
A separate set of results in \citet{jo2002}, \citet{ding2011},
\citet{jiangDing2021}, \citet{richardson2023estimating}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.

This paper expands on those latter results by presenting a new principal effect estimator for the case of one-way noncompliance, in which principal stratum membership is observed in one treatment group but not the other (future work will extend our approach to general scenarios). Our estimator boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally, estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.

We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.
Unlike any principal stratification estimator we are aware of, \geepers estimates are based on widely-used regression techniques, making them easy to compute, check, and understand for researchers without an extensive background in statistics. 
Existing methods fail to meet three needs: (1) accessibility, (2) robustness, and (3) model flexibility. \textsc{geepers} addresses all three.

%After the initial submission of this manuscript, we became aware of \citet{richardson2023estimating}, which suggests a similar, though not identical, estimator to ours, using G-estimation of a structural nested mean model \citep[e.g.][]{wallace2017r}.
%Our development of \textsc{geepers} builds on \citet{richardson2023estimating} by emphasizing connections to both parametric mixture modeling and principal score analysis, and comparing \textsc{geepers} to those methods in a simulation study.
%Moreover, \textsc{geepers}, which, in practice, is a combination of logistic regression and OLS, is more accessible to applied researchers while yielding very similar results.

%This work advances the literature in three ways: first, it makes an explicit connection between principal score methods, M-estimation, and semi-parametric estimation of principal effects using auxiliary data. Second, it provides a general estimator that relies on two of the most familiar tools in the applied education researcher's toolbox: linear and logistic regression. The \textsc{geepers} method can, therefore, make principal stratification methods accessible for a much broader range of applied education researchers. Lastly, this work includes an extensive simulation study that compares \textsc{geepers} to parametric mixture modeling and principal score weighting under a range of favorable and challenging scenarios. The simulation demonstrates the strengths and weaknesses of each method, as well as the trade-offs between them.

@@

\begin{center}
  [Table \ref{tab:factor} around here]\\
\end{center}

@@

\subsection{Results}\label{sec:simResults}
In this section, we evaluate estimates of $\eff1$. Results for estimates of $\eff0$, which are similar, can be found in the online appendix. 
In replication $b=1,\dots,500$, define the error of a given estimator $\heff1_b$ as $err_b=\heff1_b-\eff1$. The bias of $\heff1$, $\EE[\heff1-\eff1]$, is estimated as $\widehat{bias}=\sum_b err_b/500$, and the empirical standard error is defined as $\left\{\sum_b(\heff1_b-\overline{\heff1})^2/499\right\}^{1/2}$, the sample standard deviation of $\heff1$. We also report the empirical coverage of confidence intervals, the proportion of replications in which $\eff1$ is within the estimated confidence bounds of $\heff1$. Results on root-mean-squared-error %$\left(\widehat{bias}^2+SE_{emp}^2\right)^{1/2}$ 
$\left(\sum_b err_b^2/500\right)^{1/2}$
are in the online appendix.

@@

\begin{center}
  [Figure \ref{fig:alphan} around here]\\
\end{center}

@@

Our next set of results compares the performance of \geepers, \pmm,
and \psw when one or more assumptions of each method may be violated.
We present results for $n=500$ here; results from when $n=1,000$,
which are similar, can be found in the online appendix.

\begin{center}
  [Figure \ref{fig:boxplots} around here]\\
\end{center}

@@

\begin{center}
  [Table \ref{tab:coverage} around here]\\
\end{center}

@@

\begin{center}
  [Figure \ref{fig:opt} around here]\\
\end{center}

@@

Another arguable disadvantage of the observational study approach is that it overlooks the broader context of the experiment, focusing solely on whether bottom-out hints affect posttest scores rather than examining their role in the overall effectiveness of ASSISTments.
A second approach, mediation analysis, contextualizes the bottom-out effect within the larger experiment.
It, too, relies on the same non-confounding assumption as an observational study, and also suffers from an additional challenge: mediation analysis relies on a ``common support'' condition that ``the mediator state must not be a deterministic function of the treatment'' \citep[p. 225]{celli2022causal}.
However, bottom-out hints are unavailable in the other three arms of the study, violating this condition (\citealt{sales2021student} offer a solution, which we find unconvincing).

@@

\begin{center}
  [Figure \ref{fig:effects} around here]\\
\end{center}

@@

\begin{center}
  [Figure \ref{fig:compare} around here]\\
\end{center}

@@

\section{Discussion}\label{sec:discussion}
\textsc{geepers} is a straightforward approach to principal effect estimation under strong monotonicity, which is more robust---though sometimes less precise---than alternative approaches under a wide array of scenarios. It is built on widely-used regression models, making it accessible to applied researchers.

@@

\section*{AI Disclosure}
We used Grammarly Pro (Version 14.1243.0) and ChatGPT (GPT-4-turbo, current as of July 2025) for editing and language improvement.

@@

\FloatBarrier
\section*{Tables and Figures}
\FloatBarrier

\begin{table}[t]
    \caption{\label{tab:factor} Manipulated factors in simulation study}
  \centering
  %\rowcolors{1}{}{lightgray}
\begin{tabular}{r|ll}%{*{3}{c}}
  \hline
  Factor &\multicolumn{2}{l}{Levels}\\
  \hline
Sample size per condition $n$ &\multicolumn{2}{l}{$100^*,200^*,300^*,400^*,500,600^*,700^*,800^*,900^*,1000$}\\
$\alpha$ &\multicolumn{2}{l}{$\rd{0},0.2,0.3^\dagger,0.4^\dagger,0.5,0.6^\dagger,0.7^\dagger,0.8^\dagger,0.9^\dagger,1^\dagger$}\\
The distribution of $\epsilon_i$&\multicolumn{2}{l}{ Normal, 
\rd{Uniform}}\\
%$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
\multirow{4}{*}{Interactions in \eqref{eq:y-sim}}& None:  $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
& \rd{$\bm{x}:\st$}: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
& \rd{$\bm{x}:Z$}: $\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
& \rd{$\bm{x}:\st$ \& $\bm{x}:Z$}: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% between $\st$ and covariates & No, Yes\\
% Interactions between $Z$ and covariates& No, Yes\\
\hline
\multicolumn{3}{p{4.5in}}{\footnotesize Factor levels in red violate assumptions of \geepers, \pmm, or both. Factors fully crossed except where indicated.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $*$ Only tested with $\alpha=0.5$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $\dagger$ Only tested with $n=500$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}

\begin{figure}[!ht]
  \centering
  \includegraphics[width=5.5in,clip]{../simFigs/boxplotsNew.pdf}
  \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
  \label{fig:boxplots}
\end{figure}
\clearpage
